Flute and kink instabilities in a dynamically twisted flux tube with anisotropic plasma viscosity

### 2022-02-17 paper.tex
proofreading - a number of small changes 2

\begin{figure*}
  \centering
    \begin{subfigure}{0.32\textwidth}
      \includegraphics[width=\linewidth]{swi-3_pressure_12.pdf}
      \caption{$t=24$}
      \label{fig:swi-3_pressure_12}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.32\textwidth}
      \includegraphics[width=\linewidth]{swi-3_pressure_14.pdf}
      \caption{$t=28$}
      \label{fig:swi-3_pressure_14}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.32\textwidth}
      \includegraphics[width=\linewidth]{swi-3_pressure_15.pdf}
      \caption{$t=30$}
      \label{fig:swi-3_pressure_15}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.32\textwidth}
      \includegraphics[width=\linewidth]{swi-3_pressure_16.pdf}
      \caption{$t=32$}
      \label{fig:swi-3_pressure_16}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.32\textwidth}
      \includegraphics[width=\linewidth]{swi-3_pressure_17.pdf}
      \caption{$t=34$}
      \label{fig:swi-3_pressure_17}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.32\textwidth}
      \includegraphics[width=\linewidth]{swi-3_pressure_18.pdf}
      \caption{$t=36$}
      \label{fig:swi-3_pressure_18}
    \end{subfigure}
\mycaption{\rev{Gas pressure} profiles at $z=0$ during the development of the flute and
kink instabilities in the higher resistivity anisotropic case.}{The viscosity
model is anisotropic and $\eta = 10^{-3}$. In contrast to the case of
$\eta=10^{-4}$, the nonlinear development of the flute instability has time to
mix the interior of the flux tube before the onset of the kink instability, the
growth of which is \rev{likely} affected by the mixed plasma.}
\label{fig:kink_pressure_slices-3}%
\end{figure*}

Figure \ref{fig:kink_pressure_slices-3} shows a prolonged development of the
flute instability and a slow nonlinear development of the kink instability at
the higher resistivity value $\eta=10^{-3}$ in the case of anisotropic
viscosity. Due to the enhanced Ohmic heating at $\eta=10^{-3}$, the \rev{gas} pressure
gradient is substantially stronger than at $\eta=10^{-4}$ and the flute
instability is excited much earlier. Compared to the $\eta=10^{-4}$ cases, the
instability occurs further from the axis, at $r\approx0.16$, and the larger
\rev{gas} pressure gradient drives the bulges in profile further from the axis during the
nonlinear phase (Figure~\ref{fig:swi-3_pressure_12}). These bulges continue to
extend outwards and mix the plasma as they develop. The kink instability can be
observed moving the axis of the tube diagonally upwards and to the right in
Figure~\ref{fig:swi-3_pressure_15}. At this time in the $\eta=10^{-4}$ cases,
the nonlinear development of the kink was at a later stage of its development
(Figure~\ref{fig:swi-4_pressure_15}). The development of the kink then proceeds
slowly as it moves the axis of the tube through the mixed region to eventually
begin the reconnection process with the outer region of field that is typical
of the instability in this kind of flux tube as was observed by \revcite{\citet{quinnEffectAnisotropicViscosity2020}}.

@@

It is evident from the kinetic energy profile that the flute instability
develops much earlier than in the $\eta=10^{-4}$ cases and grows at an
increased rate of $\gamma = 1.06$ (Figure~\ref{fig:kink_ke_log-3}). The kink
instability grows at a rate of $\lambda \approx 0.15$, much slower than that
observed in the $\eta=10^{-4}$ cases, and much lower than the flute
instability. \rev{The time period between $t\approx28$ and $t\approx32$ is
identified as the linear stage of the kink instability by investigating the
development shown in Figure~\ref{fig:swi-3_pressure_12}. This is broadly
consistent with the tube surpassing a critical twist of
$2.59\pi$~\citep{hood1981,Torok2003} between $t=30$ and $t=35$.} One key
observation is that, despite the early and disruptive growth of the flute
instability, the kink instability still generates the bulk of the kinetic
energy (Figure~\ref{fig:kink_ke-3}).

@@

Table~\ref{tab:kink_fluting_params} summarises the quantitative differences
between the results for the two values of the resistivity $\eta$ and two models
of viscosity. \rev{The mean across the radius of the ideal estimate of the
growth rate~\eqref{eq:fluting_growth_rate2} provides the theoretical average
growth rate, which is in decent agreement with the observed rate in each case,
particularly in the more ideal case where $\eta=10^{-4}$. \rev{The
  discrepancy between predicted and observed growth rates is due to,
  in part, the growth rate estimate~\eqref{eq:fluting_growth_rate2}
  being derived under the assumption of asymptotically large values of
  $m \gg 1$, which is not true for the observed perturbation where
  $m=4$. Despite this, the predicted growth rate remains of a similar
  order of magnitude to the observed rate.} The location of the 
peak growth rate provides a prediction of where the instability will initially
grow. This radius is used in conjunction with the resonance
equation~\eqref{eq:resonant_surface} (with $m=4$) to predict the axial
wavenumber of the perturbation with the greatest linear growth, i.e. the
perturbation most likely to be observed. Again, these are in good agreement
with the observed fluting wavenumbers, which are measured at times just prior
to the accelerated development of the flute instability, that is at $t=22$ when
$\eta=10^{-3}$ and $t=26$ when $\eta = 10^{-4}$. The kink wavenumber is
measured at $t=26$ in both cases. Overall, the agreement between predicted and observed growth rates and perturbation wavenumbers allows us to conclude that the observed instability is the ideal flute instability and that the growth rate~\eqref{eq:fluting_growth_rate2} can be effectively applied to the instability within coronal loops.}

@@

\rev{Perturbing a magnetic flux tube by dynamic twisting of the flow
at the cylinder bases leads to excitation of the flute instability in
addition to the well studied kink instability.}

\rev{Our aim in performing the reported numerical experiments was to explore
the flute instability in a dynamically twisted coronal flux tube, specifically
focussing on the effect of anisotropic viscosity on the development of the
instability. In addition we wish to understand the effect the instability has
on the proceeding kink instability, the effect on the overall heating generated
through viscous and Ohmic dissipation, and the effect that varying resistivity
has on the development of the flute.}

@@

\rev{The growth rate of the kink instability is, counter-intuitively, lower in the
weakly dissipative anisotropic cases, compared to the strongly dissipative
isotropic cases, i.e. one would expect the kink instability to grow more
quickly (or at least be unaffected) when using anisotropic viscosity. Indeed,
the simulations reported by~\citet{quinnEffectAnisotropicViscosity2020} display
this behaviour, where the kink instability grows \newold{}{(marginally)} faster in the
anisotropic cases. We speculate that it is the presence of the flute modes which
negatively affects the growth of the kink instability.} {It seems unlikely that
in the linear regime the perturbations of the flute and kink are able to
directly couple, given that the kink instability generally presents at the axis
of a flux tube and the flute at some resonant surface away from the
axis.} \rev{We believe that, instead of a linear coupling, the linear kink instability is
disturbed by the more complex magnetic field configuration that arises due to
the mixing caused by the nonlinear development of the flute modes. The
complexity in the field can be seen by comparing
figures~\ref{fig:field_line_plots_iso} and~\ref{fig:field_line_plots_swi}.

Beyond the effect on the growth of the two observed instabilities, the two
viscosity models greatly affect both the viscous heating and Ohmic heating
rates. Anisotropic viscosity is naturally less dissipative than isotropic
viscosity and generates approximately two orders of magnitude less total
viscous heat than isotropic viscosity. This is somewhat offset by anisotropic
viscosity permitting greater kinetic energy release and enhanced mixing, which
in turn enhances Ohmic dissipation of heat through the generation of small,
strong current sheets. The overall effect is that more heat is generated when
the viscosity is anisotropic, similar to what has been observed in previous
work~\citep{quinnEffectAnisotropicViscosity2020}. In the context of
coronal heating, this is encouraging: the use of a less dissipative viscosity
model actually results in greater overall heating. How this result generalises
to more realistic coronal resistivities, and whether this result holds true for
other coronal instabilities, should be the subject of further investigation.}

\rev{It is difficult to distinguish the effect of the flute instability on the
viscous or Ohmic heating from that of the viscosity itself, particularly since
the flute instability is quickly disrupted by the kink instability. However, it
can be concluded from the plots of cumulative heat (Figures~\ref{fig:heating-4}
and~\ref{fig:heating_r-3}) that the majority of the viscous and Ohmic heat is
generated by the nonlinear development of the kink instability. There is
additionally some non-negligible Ohmic heating which is generated prior to the
onset of the kink, however this is attributed to the large-scale currents
associated with the twist in the field, rather than any currents created by the
flute instability. This leads us to conclude that the flute instability itself
has little direct impact on coronal heating, but instead can affect the heating
rate by slowing the development of the kink instability.}

It is likely that the $m=4$ azimuthal perturbation is excited due to influences
from the boundaries in the Cartesian box, for example through the interaction
of reflected fast waves generated in part by the driver. Performing a similar
experiment in a cylindrical numerical domain, or prescribing a variety of
perturbations in the Cartesian domain may reveal other, faster growing modes.
The modes may also be influenced by nonlinear coupling between the $m>1$ and
$m=1$ modes, as is found in the study of kink and flute
oscillations~\citep{terradasEffectMagneticTwist2018,rudermanNonlinearGenerationFluting2017}.

As the current distribution, which develops as the flux tube is twisted, is
similar to that found in the initial flux tube configuration
of~\cite{quinnEffectAnisotropicViscosity2020},the question arises why the
fluting instability is not observed in the latter. Although the current
distribution (and thus heating and \rev{gas} pressure distributions) in the tubes
of~\cite{quinnEffectAnisotropicViscosity2020} may support the flute
instability, the tube is initially perturbed with a motion close to an unstable
eigenmode of the kink instability, resulting in the instability growing from
$t=0$. In contrast, in the tubes studied here, such a perturbation must build
from numerical noise, allowing a secondary, fluting perturbation to also build
and become significant enough to observe. \rev{As an explorative alternative to
prescriptive perturbations, we recommend the use of numerical noise in the
study of coronal instabilities.}

>>>>

\begin{figure*}
  \centering
    \begin{subfigure}{0.32\textwidth}
      \includegraphics[width=\linewidth]{swi-3_pressure_12.pdf}
      \caption{$t=24$}
      \label{fig:swi-3_pressure_12}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.32\textwidth}
      \includegraphics[width=\linewidth]{swi-3_pressure_14.pdf}
      \caption{$t=28$}
      \label{fig:swi-3_pressure_14}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.32\textwidth}
      \includegraphics[width=\linewidth]{swi-3_pressure_15.pdf}
      \caption{$t=30$}
      \label{fig:swi-3_pressure_15}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.32\textwidth}
      \includegraphics[width=\linewidth]{swi-3_pressure_16.pdf}
      \caption{$t=32$}
      \label{fig:swi-3_pressure_16}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.32\textwidth}
      \includegraphics[width=\linewidth]{swi-3_pressure_17.pdf}
      \caption{$t=34$}
      \label{fig:swi-3_pressure_17}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.32\textwidth}
      \includegraphics[width=\linewidth]{swi-3_pressure_18.pdf}
      \caption{$t=36$}
      \label{fig:swi-3_pressure_18}
    \end{subfigure}
\mycaption{\rev{Gas pressure} profiles at $z=0$ during the development of the flute and
kink instabilities in the higher resistivity anisotropic case.}{The viscosity
model is anisotropic and $\eta = 10^{-3}$. In contrast to the case of
$\eta=10^{-4}$, the nonlinear development of the flute instability has time to
mix the interior of the flux tube before the onset of the kink instability, the
growth of which is \rev{likely} affected by mixing of the plasma.}
\label{fig:kink_pressure_slices-3}%
\end{figure*}

Figure \ref{fig:kink_pressure_slices-3} shows a prolonged development of the
flute instability and a slow nonlinear development of the kink instability at
the higher resistivity value $\eta=10^{-3}$ in the case of anisotropic
viscosity. Due to the enhanced Ohmic heating at $\eta=10^{-3}$, the \rev{gas} pressure
gradient is substantially stronger than at $\eta=10^{-4}$ and the flute
instability is excited much earlier. Compared to the $\eta=10^{-4}$ cases, the
instability occurs further from the axis, at $r\approx0.16$, and the larger
\rev{gas} pressure gradient drives the bulges in profile further from the axis during the
nonlinear phase (Figure~\ref{fig:swi-3_pressure_12}). These bulges continue to
extend outwards and mix the plasma as they develop. The kink instability can be
observed displacing the axis of the tube diagonally upwards and to the right in
Figure~\ref{fig:swi-3_pressure_15}. At this time in the $\eta=10^{-4}$ cases,
the nonlinear development of the kink was at a later stage of its development
(Figure~\ref{fig:swi-4_pressure_15}). The development of the kink then proceeds
slowly as it moves the axis of the tube further through the mixed
region to eventually begin the reconnection process with the outer
region of field that is typical of the instability in this kind of
flux tube as was observed by
\revcite{\citet{quinnEffectAnisotropicViscosity2020}}.

@@

It is evident from the kinetic energy profile that the flute instability
develops much earlier than in the $\eta=10^{-4}$ cases and grows at an
increased rate of $\gamma = 1.06$ (Figure~\ref{fig:kink_ke_log-3}). The kink
instability grows at a rate of $\lambda \approx 0.15$, much slower than that
observed in the $\eta=10^{-4}$ cases, and much lower than the flute
instability. \rev{The time period between $t\approx28$ and $t\approx32$ is
identified as the linear stage of the kink instability by inspecting the
development shown in Figure~\ref{fig:swi-3_pressure_12}. This is broadly
consistent with the tube surpassing a critical twist of
$2.59\pi$~\citep{hood1981,Torok2003} between $t=30$ and $t=35$.} One key
observation is that, despite the early and disruptive growth of the flute
instability, the kink instability still generates the bulk of the kinetic
energy (Figure~\ref{fig:kink_ke-3}).

@@

Table~\ref{tab:kink_fluting_params} summarises the quantitative differences
between the results for \rev{the two models
  of viscosity and} the two values of the resistivity. \rev{The
  theoretical average growth rate is computed as the mean across the
  radius of the ideal estimate~\eqref{eq:fluting_growth_rate2} and is
  in good agreement with the observed rate in each case, particularly
  in the less resistive case $\eta=10^{-4}$ which better represents
  ideal plasma. The
  discrepancy between predicted and observed growth rates is due,
  in part, to the growth rate estimate~\eqref{eq:fluting_growth_rate2}
  being derived under the assumption of asymptotically large values of
  $m \gg 1$, while the observed mode has a finite value of
  $m=4$. Despite this, the predicted growth rate is of a similar 
  magnitude to the observed rate. The location $r_s$ of the 
peak growth rate provides a prediction of where the instability will initially
grow. This radius is used in conjunction with the resonance
equation~\eqref{eq:resonant_surface}, with $m=4$, to predict the axial
wavenumber of the mode with the greatest linear growth, i.e. the
mode most likely to be observed. Again, these are in good agreement
with the observed fluting wavenumbers, which are measured at times just prior
to the accelerated development of the flute instability, that is at $t=22$ when
$\eta=10^{-3}$ and $t=26$ when $\eta = 10^{-4}$. The kink wavenumber is
measured at $t=26$ in both cases. Overall, the agreement between
predicted and observed growth rates and mode wavenumbers
allows us to conclude that the observed instability is the ideal flute
instability and that expression~\eqref{eq:fluting_growth_rate2}
can be effectively applied to estimate the growth rate of the flute
instability within coronal loops.}

@@

\rev{Perturbing a magnetic flux tube by dynamic twisting of the flow
at the cylinder bases leads to excitation of the flute instability in
addition to the well studied kink instability.}
\rev{Our aim in performing the reported numerical experiments was to explore
the flute instability in a dynamically twisted coronal flux tube, specifically
focussing on the effect of anisotropic viscosity on the development of the
instability. In addition, we wish to understand the effect the instability has
on the proceeding kink instability, the effect on the overall heating generated
through viscous and Ohmic dissipation, and the effect that varying resistivity
has on the development of the flute. Our findings are discussed below.}

@@

\rev{Counter-intuitively, the growth rate of the kink instability is lower in the
weakly dissipative anisotropic cases, compared to the strongly dissipative
isotropic cases as one would expect the kink instability to grow more
quickly (or at least be unaffected) when using anisotropic viscosity. Indeed,
the simulations reported by~\citet{quinnEffectAnisotropicViscosity2020} display
this behaviour, where the kink instability grows \newold{}{(marginally)} faster in the
anisotropic cases. We speculate that it is the presence of the flute modes which
negatively affects the growth of the kink instability.} {It seems unlikely that
in the linear regime the flute and kink modes are able to
directly couple, given that the kink instability generally presents at the axis
of a flux tube and the flute at some resonant surface away from the
axis.} \rev{We believe that, instead of a direct coupling, the linear kink instability is
disturbed by the more complex magnetic field configuration that arises due to
the mixing caused by the nonlinear development of the flute modes. The
complexity in the field can be seen by comparing
figures~\ref{fig:field_line_plots_iso} and~\ref{fig:field_line_plots_swi}.

Beyond the effect on the growth of the two observed instabilities, the two
viscosity models greatly affect both the viscous heating and Ohmic heating
rates illustrated in Figures \ref{fig:heating-4}
and~\ref{fig:heating_r-3}. Anisotropic viscosity is naturally less
dissipative than isotropic 
viscosity and generates approximately two orders of magnitude less total
viscous heat than isotropic viscosity. This is somewhat offset by anisotropic
viscosity permitting greater kinetic energy release and enhanced mixing, which
in turn enhances Ohmic dissipation of heat through the generation of
strong localised current sheets. The overall effect is that more heat is generated when
the viscosity is anisotropic, similar to what has been observed in previous
work~\citep{quinnEffectAnisotropicViscosity2020}. In the context of
coronal heating, this is encouraging: the use of a less dissipative viscosity
model actually results in greater overall heating. How this finding generalises
to more realistic coronal resistivities, and whether it holds true for
other coronal instabilities, should be the subject of further investigations.}

\rev{It is difficult to distinguish the effect of the flute instability on the
viscous or Ohmic heating from that of the viscosity itself, particularly since
the flute instability is quickly disrupted by the kink instability. However, it
can be concluded from the plots of cumulative heat (Figures~\ref{fig:heating-4}
and~\ref{fig:heating_r-3}) that the bulk of the viscous and Ohmic heat is
generated in the nonlinear phase of the kink instability. There is
additionally some non-negligible Ohmic heating generated prior to the
onset of the kink, however this is attributed to the large-scale currents
associated with the twist in the field, rather than any currents created by the
flute instability. This leads us to conclude that the flute instability itself
has little direct impact on coronal heating, but instead can affect the heating
rate by slowing the development of the kink instability.}

It is likely that the $m=4$ azimuthal mode is excited due to influences
from the boundaries in the Cartesian box, for example through the interaction
of reflected fast waves generated in part by the driver. Performing a similar
experiment in a cylindrical numerical domain, or prescribing a variety of
perturbations in the Cartesian domain may reveal other, faster growing modes.
The modes may also be influenced by nonlinear coupling between the $m>1$ and
$m=1$ modes, as is found in the study of kink and flute
oscillations~\citep{terradasEffectMagneticTwist2018,rudermanNonlinearGenerationFluting2017}.

As the current distribution, which develops as the flux tube is twisted, is
similar to that found in the initial flux tube configuration
of~\cite{quinnEffectAnisotropicViscosity2020}, the question arises why the
fluting instability is not observed in the latter. Although the current
distribution (and thus heating and pressure distributions) in the tubes
of~\cite{quinnEffectAnisotropicViscosity2020} may support the flute
instability, the tube is initially perturbed with a motion close to an unstable
eigenmode of the kink instability, resulting in the instability
growing immediately from $t=0$. In contrast, in the tubes studied here, such a perturbation must grow
from numerical noise, allowing a secondary, fluting modes perturbation to also develop
and become significant enough to observe. \rev{As an explorative alternative to
prescriptive perturbations, we recommend the use of numerical noise in the
study of coronal instabilities.}
